---
title: "NBA Games Historical Analysis & Scorigami Poster"
author: "Antigravity Pair Programmer"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
---

## Introduction

This report analyzes historical NBA game data loaded from `Games.csv` (derived from the [Kaggle NBA Dataset: Box Scores and Stats](https://www.kaggle.com/datasets/eoinamoore/historical-nba-data-and-player-box-scores)). The dataset covers games from 1949 to the present, capturing team scores, game dates, locations, attendance, and matchups.

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(knitr)
library(scales)

# Read the dataset
games <- read_csv("Games.csv", show_col_types = FALSE)
```

## Dataset Characteristics

Let's look at the dimensions of the raw dataset:

```{r dimensions}
dim_info <- tibble(
  Metric = c("Total Games (Rows)", "Total Attributes (Columns)"),
  Value = c(nrow(games), ncol(games))
)
kable(dim_info, caption = "Dataset Dimensions")
```

### Column Specifications

Here is a summary of the columns present in the dataset:

```{r columns}
cols_summary <- tibble(
  ColumnName = names(games),
  DataType = sapply(games, function(x) class(x)[1]),
  SampleValue = sapply(games, function(x) as.character(x[1]))
)
kable(cols_summary, caption = "Column Meta-information")
```

```{r cleaning, message=FALSE, warning=FALSE}
# Clean and process the dataset, filtering out zero-score anomalies
games_clean <- games %>%
  mutate(
    gameDate = as.Date(gameDate),
    homeScore = as.numeric(homeScore),
    awayScore = as.numeric(awayScore),
    winner = as.character(winner),
    hometeamId = as.character(hometeamId),
    awayteamId = as.character(awayteamId),
    Year = year(gameDate)
  ) %>%
  filter(!is.na(homeScore) & !is.na(awayScore) & !is.na(gameDate) & homeScore > 0 & awayScore > 0) %>%
  mutate(
    home_win = ifelse(winner == hometeamId, 1, 0),
    total_score = homeScore + awayScore,
    score_diff = homeScore - awayScore
  )
```

## NBA Historical Scoring Trends (1949 - Present)

How has scoring evolved in the NBA over the decades? The plot below highlights the average combined score of games per year.

```{r scoring-trend, fig.width=10, fig.height=6}
scoring_trend <- games_clean %>%
  group_by(Year) %>%
  summarise(
    Avg_Total_Points = mean(total_score, na.rm = TRUE),
    Avg_Home_Score = mean(homeScore, na.rm = TRUE),
    Avg_Away_Score = mean(awayScore, na.rm = TRUE),
    Game_Count = n()
  )

ggplot(scoring_trend, aes(x = Year)) +
  geom_line(aes(y = Avg_Total_Points, color = "Total Combined Points"), size = 1.2) +
  geom_line(aes(y = Avg_Home_Score, color = "Home Team Score"), size = 0.8, linetype = "dashed") +
  geom_line(aes(y = Avg_Away_Score, color = "Away Team Score"), size = 0.8, linetype = "dashed") +
  scale_color_manual(values = c(
    "Total Combined Points" = "#2c3e50", 
    "Home Team Score" = "#e74c3c", 
    "Away Team Score" = "#3498db"
  )) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Historical NBA Scoring Trends (1949 - Present)",
    subtitle = "Evolution of average home, away, and combined points per game",
    x = "Year",
    y = "Average Points",
    color = "Metric"
  )
```

We see a clear dip in the 1990s and early 2000s (often referred to as the "dead-ball" or defensive era) followed by a sharp rise starting around 2015 with the popularity of the three-point shot.

## Home Court Advantage Over Time

Is playing at home still a strong advantage? Let's analyze the home-court win rate by year.

```{r home-advantage, fig.width=10, fig.height=6}
home_advantage <- games_clean %>%
  group_by(Year) %>%
  summarise(
    Home_Win_Pct = mean(home_win, na.rm = TRUE) * 100
  )

ggplot(home_advantage, aes(x = Year, y = Home_Win_Pct)) +
  geom_line(color = "#16a085", size = 1.2) +
  geom_point(color = "#117a65", size = 2) +
  geom_hline(yintercept = 50, linetype = "dotted", color = "red", size = 0.8) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Home Court Win Percentage Over Time",
    subtitle = "Percentage of games won by the home team (50% indicates no advantage)",
    x = "Year",
    y = "Win Percentage (%)"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1))
```

Historically, home-court win rates hovered around 60-70% in the early years but have steadily converged closer to 55-58% in recent seasons.

## Top 10 Highest-Scoring Matches

Let's list the top 10 highest-scoring matches in NBA history contained within this dataset:

```{r highest-scoring}
top_10 <- games_clean %>%
  arrange(desc(total_score)) %>%
  slice(1:10) %>%
  select(gameDate, gameType, hometeamCity, hometeamName, homeScore, awayteamCity, awayteamName, awayScore, total_score)

kable(top_10, col.names = c("Date", "Type", "Home City", "Home Team", "Home Score", "Away City", "Away Team", "Away Score", "Total Score"), caption = "Top 10 Highest-Scoring NBA Games")
```

## Top 10 Arenas by Average Attendance

Which arenas have the highest average attendance historically (filtering for games that recorded attendance)?

```{r attendance}
top_arenas <- games_clean %>%
  filter(!is.na(attendance) & attendance > 5000) %>%
  group_by(arenaName, arenaCity, arenaState) %>%
  summarise(
    Games_Count = n(),
    Avg_Attendance = mean(attendance),
    Max_Attendance = max(attendance),
    .groups = "drop"
  ) %>%
  filter(Games_Count >= 20) %>%
  arrange(desc(Avg_Attendance)) %>%
  slice(1:10)

kable(top_arenas, col.names = c("Arena Name", "City", "State", "Games Recorded", "Avg Attendance", "Max Attendance"), caption = "Top 10 Arenas by Average Attendance (min. 20 games)")
```

---

## NBA Scorigami Poster Layout

The section below replicates the visual hierarchy and style of the **NBA Scorigami** reference poster (`test_scores.png`). It computes all stats dynamically and renders the two historical timeline insets and the main annotated Scorigami grid.

```{r scorigami-calculations, message=FALSE, warning=FALSE}
# Setup boundaries
min_s <- 15
max_s <- 192

# Calculate unique score counts and first occurrences
games_scores <- games_clean %>%
  filter(homeScore >= min_s & homeScore <= max_s & awayScore >= min_s & awayScore <= max_s)

# Home/Visitor Scorigami Grid Data
home_visitor_scores <- games_scores %>%
  group_by(homeScore, awayScore) %>%
  summarise(
    Count = n(),
    First_Date = min(gameDate),
    First_Home = first(hometeamName),
    First_Away = first(awayteamName),
    .groups = "drop"
  )

# Calculate statistics over time for the mini-charts
games_timeline <- games_scores %>%
  mutate(
    w = pmax(homeScore, awayScore),
    l = pmin(homeScore, awayScore),
    score_key = paste(w, l, sep = "-")
  )

score_firsts <- games_timeline %>%
  group_by(score_key) %>%
  summarise(first_date = min(gameDate), .groups = "drop") %>%
  mutate(first_year = year(first_date))

new_scores_per_year <- score_firsts %>%
  group_by(first_year) %>%
  summarise(new_scores_count = n(), .groups = "drop")

stats_by_year <- tibble(Year = 1946:2026) %>%
  left_join(new_scores_per_year, by = c("Year" = "first_year")) %>%
  mutate(new_scores_count = coalesce(new_scores_count, 0)) %>%
  mutate(cumulative_scores = cumsum(new_scores_count))

# Unique scores count in home-visitor coordinates
unique_hv_count <- nrow(home_visitor_scores)
```

<!-- HTML Layout for the Poster using Pandoc Fenced Divs for styling wrapper blocks -->

::: {style="background-color: #eae7e1; padding: 2rem; border-radius: 12px; border: 1px solid #d1cfc9; font-family: 'Outfit', sans-serif; color: #1a2035; margin-top: 2rem; box-shadow: 0 4px 20px rgba(0,0,0,0.06);"}
  
<!-- Header Row -->
::: {style="display: flex; justify-content: space-between; align-items: flex-start; border-bottom: 2px solid #d1cfc9; padding-bottom: 1.25rem; margin-bottom: 1.5rem;"}
<div>
  <h1 style="font-size: 3.2rem; font-weight: 800; margin: 0; color: #1a2035; letter-spacing: -0.04em; font-family: 'Outfit', sans-serif; border: none; padding: 0;">NBA Scorigami</h1>
  <p style="font-size: 1rem; color: #4b5563; max-width: 650px; margin-top: 0.5rem; line-height: 1.45; font-family: 'Inter', sans-serif;">
    <strong>Scorigami</strong> in sport is a final score that has never happened before in a sport or league's history. 
    The poster below shows all unique scores in the NBA from 1946 to 2026.
  </p>
</div>
<div style="text-align: right; font-size: 0.8rem; color: #6b7280; line-height: 1.5; font-family: 'Inter', sans-serif;">
  <strong>DESIGN REFERENCE:</strong> Alex Varlamov<br>
  <strong>DATA:</strong> kaggle.com (Games.csv)<br>
  <strong>REPLICATED BY:</strong> Antigravity Pair Programmer
</div>
:::

<!-- Insets and Stats Area -->
::: {style="display: grid; grid-template-columns: 1fr 2fr; gap: 2rem; margin-bottom: 2rem;"}

<!-- Unique Scores Stats Card -->
::: {style="background-color: #ffffff; border: 1px solid #d1cfc9; border-radius: 8px; padding: 1.5rem; display: flex; flex-direction: column; justify-content: center; align-items: center; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.03);"}
<span style="font-size: 0.85rem; font-weight: 600; text-transform: uppercase; letter-spacing: 0.08em; color: #6b7280; font-family: 'Inter', sans-serif;">Unique Scores (H vs. V)</span>
<span style="font-size: 3.4rem; font-weight: 800; color: #1e1b4b; margin: 0.5rem 0; font-family: 'Outfit', sans-serif;">`r unique_hv_count`</span>
<span style="font-size: 0.85rem; color: #4b5563; font-weight: 500; font-family: 'Inter', sans-serif;">Years: 1946 - 2026</span>
:::
    
<!-- Mini Charts Container -->
::: {style="display: grid; grid-template-columns: 1fr 1fr; gap: 1.5rem;"}

<!-- Mini Chart 1: Unique Scores per Year -->
::: {style="background-color: #ffffff; border: 1px solid #d1cfc9; border-radius: 8px; padding: 1rem; box-shadow: 0 1px 3px rgba(0,0,0,0.03);"}
<h4 style="font-size: 0.75rem; margin-bottom: 0.5rem; text-transform: uppercase; color: #4b5563; font-weight: 700; font-family: 'Inter', sans-serif; letter-spacing: 0.05em;">Unique Scores per Year</h4>

```{r mini-chart-new, fig.width=4.5, fig.height=1.8, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(stats_by_year, aes(x = Year, y = new_scores_count)) +
  geom_area(fill = "#8b949e", color = "#475569", alpha = 0.85, size = 0.5) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text.y = element_text(size = 7, color = "#64748b"),
    axis.text.x = element_text(size = 7, color = "#64748b")
  ) +
  scale_x_continuous(breaks = c(1950, 1975, 2000, 2026))
```
:::
      
<!-- Mini Chart 2: Cumulative Unique Scores -->
::: {style="background-color: #ffffff; border: 1px solid #d1cfc9; border-radius: 8px; padding: 1rem; box-shadow: 0 1px 3px rgba(0,0,0,0.03);"}
<h4 style="font-size: 0.75rem; margin-bottom: 0.5rem; text-transform: uppercase; color: #4b5563; font-weight: 700; font-family: 'Inter', sans-serif; letter-spacing: 0.05em;">Cumulative Unique Scores</h4>

```{r mini-chart-cumulative, fig.width=4.5, fig.height=1.8, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(stats_by_year, aes(x = Year, y = cumulative_scores)) +
  geom_area(fill = "#8b949e", color = "#475569", alpha = 0.85, size = 0.5) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    axis.text.y = element_text(size = 7, color = "#64748b"),
    axis.text.x = element_text(size = 7, color = "#64748b")
  ) +
  scale_x_continuous(breaks = c(1950, 1975, 2000, 2026)) +
  scale_y_continuous(labels = comma_format())
```
:::

:::

:::
  
<!-- Main Grid Plot Container -->
::: {style="background-color: #ffffff; border: 1px solid #d1cfc9; border-radius: 10px; padding: 1.5rem; box-shadow: 0 1px 4px rgba(0,0,0,0.03); display: flex; justify-content: center; align-items: center; overflow: hidden;"}

```{r main-scorigami-plot, fig.width=11.5, fig.height=11.5, echo=FALSE, message=FALSE, warning=FALSE}
# Create complete grid space
all_cells <- expand.grid(
  homeScore = min_s:max_s,
  awayScore = min_s:max_s
)

# Split occurred scores to style single occurrences differently
scores_once <- home_visitor_scores %>% filter(Count == 1)
scores_many <- home_visitor_scores %>% filter(Count > 1)

# Setup manual annotations for extreme games on both sides of the diagonal
# X corresponds to home score, Y corresponds to visitor/away score
annotations_df <- tibble(
  x = c(184, 175, 171, 161, 158, 18, 46, 54, 59, 82, 84, 48, 48, 48, 120, 130, 108, 104, 102, 78, 72, 78, 78, 72),
  y = c(186, 176, 166, 168, 162, 19, 48, 46, 48, 48, 46, 42, 44, 38, 156, 150, 154, 152, 144, 126, 132, 124, 112, 118),
  label = c("DEN - DET", "LAC - SAC", "SAS - MIL", "ATL - CHI", "DEN - GSW", "MNL - FTW", "BOS - NYK", "PHW - BOS", "CHS - NYK", "NYK - TRH", "NYK - PHW", "WSC - PIT", "BOS - PIT", "BOS - WSC", "POR - GSW", "GSW - DAL", "SAC - UTA", "PHI - NYK", "TOR - NYK", "NYK - LAL", "CLE - BOS", "GSW - LAL", "NYK - UTA", "BOS - DET"),
  nudge_x = c(6, 6, 6, 6, 6, -7, -7, -6, -5, 6, 6, -6, -6, -6, -6, -6, -6, -6, -6, -6, -6, -6, -6, -6),
  nudge_y = c(6, 6, -6, 6, -6, 6, -6, -6, 5, 5, -5, -5, 5, -5, 6, 6, 6, 6, 6, 6, 6, -6, 6, -6)
)

# Custom Theme Matching Poster Style
theme_poster <- theme_minimal(base_family = "sans") +
  theme(
    plot.background = element_rect(fill = "#fbfaf7", color = NA),
    panel.background = element_rect(fill = "#fbfaf7", color = NA),
    panel.grid.major = element_blank(), # Drawn manually below
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "bold", size = 11, color = "#1a2035"),
    axis.text = element_text(size = 9, color = "#4b5563"),
    plot.title = element_text(face = "bold", size = 18, color = "#1a2035", hjust = 0.5),
    plot.subtitle = element_text(size = 11, color = "#4b5563", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 9, color = "#1a2035"),
    legend.text = element_text(size = 8, color = "#4b5563"),
    aspect.ratio = 1
  )

ggplot() +
  # 1. Draw background grid borders (unoccurred grid tiles)
  geom_tile(data = all_cells, aes(x = homeScore, y = awayScore), fill = "#ffffff", color = "#eaeaea", size = 0.05) +
  # 2. Overlay single occurrences (black/dark grey squares)
  geom_tile(data = scores_once, aes(x = homeScore, y = awayScore), fill = "#18181b", color = "#eaeaea", size = 0.05) +
  # 3. Overlay multiple occurrences (glowing plasma heatmap scale)
  geom_tile(data = scores_many, aes(x = homeScore, y = awayScore, fill = Count), color = "#eaeaea", size = 0.05) +
  # Heatmap color gradient (Plasma scale)
  scale_fill_viridis_c(
    option = "plasma",
    trans = "log10",
    labels = comma_format(),
    name = "Occurrences\n(Log Scale)"
  ) +
  # 4. Draw major grid lines (every 10 points) on top of the tiles
  annotate(
    "segment", 
    x = min_s, xend = max_s, 
    y = seq(20, 190, 10), yend = seq(20, 190, 10), 
    color = "#d1cfc9", size = 0.25
  ) +
  annotate(
    "segment", 
    x = seq(20, 190, 10), xend = seq(20, 190, 10), 
    y = min_s, yend = max_s, 
    color = "#d1cfc9", size = 0.25
  ) +
  # 5. Dash diagonal tie-line reference
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50", alpha = 0.6, size = 0.45) +
  # 6. Line connectors for annotations
  geom_segment(
    data = annotations_df,
    aes(x = x, y = y, xend = x + nudge_x * 0.75, yend = y + nudge_y * 0.75),
    color = "#1f2937",
    size = 0.35
  ) +
  # 7. Text labels for annotations
  geom_text(
    data = annotations_df,
    aes(x = x + nudge_x, y = y + nudge_y, label = label),
    size = 2.8,
    color = "#111827",
    fontface = "bold",
    family = "sans"
  ) +
  theme_poster +
  labs(
    title = "Home PTS (X) vs. Visitor PTS (Y)",
    subtitle = "Replicating the original poster grid design with exact guidelines and labels",
    x = "Home Points",
    y = "Visitor Points"
  ) +
  scale_x_continuous(breaks = seq(20, 180, 20), limits = c(15, 192)) +
  scale_y_continuous(breaks = seq(20, 180, 20), limits = c(15, 192))
```
:::

:::
