---
title: "NBA Games Historical Analysis"
author: "Antigravity Pair Programmer"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
---

## Introduction

This report analyzes historical NBA game data loaded from `Games.csv` (derived from the [Kaggle NBA Dataset: Box Scores and Stats](https://www.kaggle.com/datasets/eoinamoore/historical-nba-data-and-player-box-scores)). The dataset covers games from 1949 to the present, capturing team scores, game dates, locations, attendance, and matchups.

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(knitr)
library(scales)

# Read the dataset
games <- read_csv("Games.csv", show_col_types = FALSE)
```

## Dataset Characteristics

Let's look at the dimensions of the raw dataset:

```{r dimensions}
dim_info <- tibble(
  Metric = c("Total Games (Rows)", "Total Attributes (Columns)"),
  Value = c(nrow(games), ncol(games))
)
kable(dim_info, caption = "Dataset Dimensions")
```

### Column Specifications

Here is a summary of the columns present in the dataset:

```{r columns}
cols_summary <- tibble(
  ColumnName = names(games),
  DataType = sapply(games, function(x) class(x)[1]),
  SampleValue = sapply(games, function(x) as.character(x[1]))
)
kable(cols_summary, caption = "Column Meta-information")
```

```{r cleaning, message=FALSE, warning=FALSE}
# Clean and process the dataset
games_clean <- games %>%
  mutate(
    gameDate = as.Date(gameDate),
    homeScore = as.numeric(homeScore),
    awayScore = as.numeric(awayScore),
    winner = as.character(winner),
    hometeamId = as.character(hometeamId),
    awayteamId = as.character(awayteamId),
    Year = year(gameDate)
  ) %>%
  filter(!is.na(homeScore) & !is.na(awayScore) & !is.na(gameDate)) %>%
  mutate(
    home_win = ifelse(winner == hometeamId, 1, 0),
    total_score = homeScore + awayScore,
    score_diff = homeScore - awayScore
  )
```

## NBA Historical Scoring Trends (1949 - Present)

How has scoring evolved in the NBA over the decades? The plot below highlights the average combined score of games per year.

```{r scoring-trend, fig.width=10, fig.height=6}
scoring_trend <- games_clean %>%
  group_by(Year) %>%
  summarise(
    Avg_Total_Points = mean(total_score, na.rm = TRUE),
    Avg_Home_Score = mean(homeScore, na.rm = TRUE),
    Avg_Away_Score = mean(awayScore, na.rm = TRUE),
    Game_Count = n()
  )

ggplot(scoring_trend, aes(x = Year)) +
  geom_line(aes(y = Avg_Total_Points, color = "Total Combined Points"), size = 1.2) +
  geom_line(aes(y = Avg_Home_Score, color = "Home Team Score"), size = 0.8, linetype = "dashed") +
  geom_line(aes(y = Avg_Away_Score, color = "Away Team Score"), size = 0.8, linetype = "dashed") +
  scale_color_manual(values = c(
    "Total Combined Points" = "#2c3e50", 
    "Home Team Score" = "#e74c3c", 
    "Away Team Score" = "#3498db"
  )) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Historical NBA Scoring Trends (1949 - Present)",
    subtitle = "Evolution of average home, away, and combined points per game",
    x = "Year",
    y = "Average Points",
    color = "Metric"
  )
```

We see a clear dip in the 1990s and early 2000s (often referred to as the "dead-ball" or defensive era) followed by a sharp rise starting around 2015 with the popularity of the three-point shot.

## Home Court Advantage Over Time

Is playing at home still a strong advantage? Let's analyze the home-court win rate by year.

```{r home-advantage, fig.width=10, fig.height=6}
home_advantage <- games_clean %>%
  group_by(Year) %>%
  summarise(
    Home_Win_Pct = mean(home_win, na.rm = TRUE) * 100
  )

ggplot(home_advantage, aes(x = Year, y = Home_Win_Pct)) +
  geom_line(color = "#16a085", size = 1.2) +
  geom_point(color = "#117a65", size = 2) +
  geom_hline(yintercept = 50, linetype = "dotted", color = "red", size = 0.8) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Home Court Win Percentage Over Time",
    subtitle = "Percentage of games won by the home team (50% indicates no advantage)",
    x = "Year",
    y = "Win Percentage (%)"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1))
```

Historically, home-court win rates hovered around 60-70% in the early years but have steadily converged closer to 55-58% in recent seasons.

## Top 10 Highest-Scoring Matches

Let's list the top 10 highest-scoring matches in NBA history contained within this dataset:

```{r highest-scoring}
top_10 <- games_clean %>%
  arrange(desc(total_score)) %>%
  slice(1:10) %>%
  select(gameDate, gameType, hometeamCity, hometeamName, homeScore, awayteamCity, awayteamName, awayScore, total_score)

kable(top_10, col.names = c("Date", "Type", "Home City", "Home Team", "Home Score", "Away City", "Away Team", "Away Score", "Total Score"), caption = "Top 10 Highest-Scoring NBA Games")
```

## Top 10 Arenas by Average Attendance

Which arenas have the highest average attendance historically (filtering for games that recorded attendance)?

```{r attendance}
top_arenas <- games_clean %>%
  filter(!is.na(attendance) & attendance > 5000) %>%
  group_by(arenaName, arenaCity, arenaState) %>%
  summarise(
    Games_Count = n(),
    Avg_Attendance = mean(attendance),
    Max_Attendance = max(attendance),
    .groups = "drop"
  ) %>%
  filter(Games_Count >= 20) %>%
  arrange(desc(Avg_Attendance)) %>%
  slice(1:10)

kable(top_arenas, col.names = c("Arena Name", "City", "State", "Games Recorded", "Avg Attendance", "Max Attendance"), caption = "Top 10 Arenas by Average Attendance (min. 20 games)")
```
